---
title: "Worldbank Internet Usage Report"
author: "Tom Wheetley"
date: "10-26-2022"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    keep_tex: true
header-includes:
  - \usepackage{array}
  - \usepackage{booktabs}
  - \usepackage{colortbl}
  - \usepackage{float}
  - \usepackage{longtable}
  - \usepackage{makecell}
  - \usepackage{multirow}
  - \usepackage[normalem]{ulem}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage{wrapfig}
  - \usepackage{xcolor}

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =FALSE,message=FALSE,warning=FALSE, cache=TRUE, cache.lazy=FALSE)
```

```{r tinytex-options, include=FALSE}
options(tinytex.verbose = TRUE)
```

```{r install-library}
# Install necessary packages if not exist

if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(ggplot2)) install.packages("ggplot2")
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(dplyr)) install.packages("dplyr", repos = "http://cran.us.r-project.org")
```

```{r load-libraries}
#Libraries

library(data.table)
library(ggplot2)
library(tidyverse)
library(caret)
library(dplyr)



############################
# Define the RMSE function that will be used
RMSE <- function(true_poverty = NULL, predicted_poverty = NULL) {
  sqrt(mean((true_poverty - predicted_poverty)^2))
}

dl <- tempfile()
download.file("https://sds-platform-private.s3-us-east-2.amazonaws.com/uploads/P2-Demographic-Data.csv", dl)
demographics <- fread(dl, select = c("Country Name",	"Country Code",	"Birth rate",	"Internet users",	"Income Group"))


dl_crv <- tempfile()
download.file("https://sds-platform-private.s3-us-east-2.amazonaws.com/uploads/P2-Country-Region-Vectors.zip", dl_crv)

source(unzip(dl_crv)) 

#Internet History
dl_ih_zip <- tempfile()
download.file("https://databank.worldbank.org/data/download/WDI_csv.zip", dl_ih_zip)
df_ih <- readLines(unzip(dl_ih_zip, "WDIData.csv"))

#Load internet usage data
#you are going to need to select the internetusage.csv here. The CSV was not available for auto download from https://ourworldindata.org/internet

internetusage <- read.csv(file.choose())

#Rename the columns
colnames(internetusage) <- c("Country", "Code", "Year", "Internet.Usage")

poverty_explorer <- read.csv(file.choose())
df_poverty <- poverty_explorer %>% select(Below..2.15.a.day,	Entity,	Year)


```

\newpage

# Introduction

This is the "Choose Your Own, (CYO), project for the Harvard X data
Science CapStone Course. I will be looking at internet usage and how
poverty can effect access to it as well as how decling birth rates are
often see in countries that are improving their economic environments
and this can lead to more internet access.

# Overview

Birth rate in general, decline as the economic picture improves and we
can see internet usage increases in these countries. The World bank in
2017 set the poverty line to \$2.15 cents per person, per day. I will be
using this mark to try to predict when a country may cross that
threshold.

# Executive Summary

During the COVID-19 pandemic, many health agencies and service providers
were forced to close their offices to prevent transmission of the virus.
However to serve their clients many providers switched to providing
medical care through tele-health services. The providers were not able
to provide all services over the phone, some required internet access.
For many developing countries access to the internet was a factor in
gaining access to critical care. Many families dealing with poverty and
the economic consequences of the pandemic were unable to access the care
or information they required. In addition, populations with worse
internet access also tend to have higher rates of chronic conditions and
worse health outcomes, suggesting that as a country develops the
reliance on internet becomes even more critical. This does not suggest
that internet access pull people out of poverty, but rather is the
result of improving economic conditions.

#Key components: Function RMSE -Root Mean Square Error (RMSE) will
be used to gauge the success of the predictions, measuring the
difference between the actual and predicted value. Two different models
will be used to evaluate RMSE, A more accurate prediction will be
identified by a lower RMSE.

The following is the calculations for RMSE:

$$ RMSE = \sqrt{\frac{1}{N}\displaystyle\sum_{u,i} (\hat{y}_{u,i}-y_{u,i})^{2}} $$


\newpage
# Data downloads - There are 5 data sources for the project 3 are
downloaded automatically, the internetusage.csv file and the
poverty_explorer.csv are included with the uploaded files and will have
to be loaded. The code is provided and the "Choose" function is used to
browse to the file. \## Data cleansing - Cleansing of the data is
performed on the data sets removing nulls and NAs removing and renaming
columns for ease of use \## Visualizing the data - Data will be
visualized with 4 plots. The first showing birth Rates and percent of
internet users by income group. The second shows Birth Rate by Region
and the percentage of people with internet access. The Third shows mean
internet usage by year, by region The forth shows the mean poverty by
region for the same time frame as plot 3 for the same regions

\newpage

# Data visualizations

The poverty, internet usage and birth rate data are combined and loaded
into two data sets: pov_exp for training purpose and valid for the
validation.

There are total of 3 columns in both pov_exp and valid datasets:
pct_below is data type num shows the percentage of people with in a
coutry living below th poverty line country is data type chr and give
the name of the country Year is the year for the sample

#Birth Rates and Internet Usage

##Birth rate is defined as the number of births per 1000 adults. The
plot shows the relationship between declining birth rates and the
percentage of the population with internet access.The higher incomees
show a greater internet usage and lower birth rate.

## PLOT ONE
PLOT ONE - Birth Rates and internet use by income group You can see the groups with 
the lower birth rate are generally in the higher income group and have a higher higher percentage of internet adoption.

```{r plot 1}
qplot(data=demographics, x=`Internet users`, y=`Birth rate`, size=I(3),
color=`Income Group`)
```

```{r Data management}
##########################
#Put vectors into Data Frame
CountriesRegionsDF <- data.frame(Countries_2012_Dataset, Codes_2012_Dataset, Regions_2012_Dataset)

#head(CountriesRegionsDF)

#Change the column names
colnames(CountriesRegionsDF) <- c("Country", "Code", "Region")



#########################
#Merge the data frame

mergeddf <- merge(demographics, CountriesRegionsDF, by.x = "Country Code", by.y = "Code")

#Remove redundant column
mergeddf$`Country Name` <- NULL


#merge Internet usage with regions
mergediuregionsdf <- merge(internetusage, CountriesRegionsDF, by.x = "Code", by.y = "Code")

#Remove redundant column
mergediuregionsdf$Country.y <-NULL

#Add in percentage of people below the poverty line $2.15 a day

df_final = merge(mergediuregionsdf, df_poverty, by.x=c("Country.x", "Year"), by.y=c("Entity", "Year"))


df_intu <- mergediuregionsdf %>%
  select(Internet.Usage, Region, Year) %>%
  group_by (Region, Year) %>%
  summarize(Mean_Internet_Usage = mean(Internet.Usage, na.rm = TRUE))

########################
#Visualize the New data Frame

options("scipen"=100, "digits"=4)
```

\newpage

## PLOT TWO
PLOT TWO - Birth Rates vs Internet usage by Continent. (Africa shows
the slowest adoption) In the plot below you can see the Africa triangles
are predominately in the upper left showing a higher birth rate than the 
higher income countries of Europe represented by the green triangles.



```{r plots 2}
qplot(data=mergeddf, x=`Internet users`,
y=`Birth rate`, size=I(3), color=Region, shape=I(17), alpha =I(0.4),
main="Birth Rate vs Internet Usage")
```

\newpage
## PLOT THREE
PLOT THREE - we can see the adoption of the internet amoung the develop countries of teh world 
well over 50% by 2020, however Africa lags behind around 37%.

```{r plots 3}
#PLOT THREE - shows Average Internet usage by Continent
ggplot(data=df_intu, aes(x=Year, y=Mean_Internet_Usage, group=Region,
color=Region)) + geom_line()
```


```{r prepare the data}

df_final_plot <- df_final %>%
  select(Below..2.15.a.day, Region, Year) %>%
  group_by (Region, Year) %>%
  summarize(Mean_Poverty_By_Region = mean(Below..2.15.a.day, na.rm = TRUE))

df_final_plot <- df_final_plot %>% 
  group_by(Region, Year) %>% 
  mutate(Mean_Poverty_By_Region = replace_na(Mean_Poverty_By_Region, min(Mean_Poverty_By_Region, na.rm = TRUE)))

any(is.na(df_final_plot))
```


\newpage

## PLOT FOUR
PLOT FOUR - show us the mean poverty by region over the last 30 years.

```{r plot 4}
ggplot(data=df_final_plot, aes(x=Year, y=Mean_Poverty_By_Region,
group=Region, color=Region)) + geom_smooth(se = FALSE, method = "loess",
formula = y ~ x)
```

\newpage
## PLOT FIVE
PLOT FIVE - shows the decline in poverty by region and the years, and can
be compared to the previous plot to show the decline in poverty and the 
increase in internet usage. As the countries develop, birth rates decline and 
adoption of technology increases.


```{r plot 5}
ggplot() + geom_smooth(data=df_final_plot,
aes(x=Year, y=Mean_Poverty_By_Region, group=Region, color=Region) ,
method = "loess", formula = y ~ x)
```

\newpage

```{r Validation and Results}
# Validation set will be 10% of poverty_explorer dataset
set.seed(1)
test_index <- createDataPartition(y = poverty_explorer$Below..2.15.a.day, times = 1, p = 0.1, list = FALSE)
pov_exp <- poverty_explorer[-test_index,]
temp <- poverty_explorer[test_index,]

validation <- temp

# Mean-Baseline Model#########################
#World bank calculates the poverty line at $2.15 per person per day 
# Calculate the mean of pov_exp dataset and assign to mu_hat

mu_hat <- mean(pov_exp$Below..2.15.a.day)

# Predict the RMSE on the validation dataset

rmse_mean_model_result <- RMSE(validation$Below..2.15.a.day, mu_hat)

# Creating a results dataframe that contains all RMSE results and add the result to it

results <- data.frame(model="Mean-Baseline Model", RMSE=rmse_mean_model_result)
# results
```

print(results)

```{r Models}

##############################
#Poverty Based Model

valid <- validation %>% select(Below..2.15.a.day, Entity, Year)
colnames(valid) <- c("pct_below", "country", "Year")


#Rename the columns
colnames(df_poverty) <- c("pct_below", "country", "Year")
colnames(df_final) <- c("country", "Year", "Code", "Internet.Usage", "Region", "pct_below")

# Calculate the average poverty by country

pov_avgs <- df_final %>%
  left_join(df_poverty, by='country') %>%
  group_by(Region) %>%
  summarize(b_u = mean( df_poverty$pct_below - mu_hat))

# Calculate the predicted poverty on validation dataset

rmse_poverty_based_model <- mean(df_final$pct_below * (df_final$Internet.Usage/100)) - mu_hat

head(rmse_poverty_based_model)

rmse_poverty_based_model_result <- RMSE(valid$pct_below, rmse_poverty_based_model)

# Adding the results to the results dataset

results <- results %>% add_row(model="Poverty Based Model", RMSE=rmse_poverty_based_model_result)
```

# Results 
  Results Section: The mean-baseline model appears to be the more accurate of the two. 
```{r results section}


print(results)
```

print(results)


# Conclusion 
  In conclusion the data shows that people in countries
climbing out of poverty are in a better economic position to access the
internet.This bring other benefits like better access to information and
healthcare. The small data sets make accurate reporting difficult, but
overall you can see the connection to improved economic situations and
the adoption of the internet.

This model could likely be further improved by including other
parameters (time of day,time of year,year, genre), or by using more
sophisticated machine learning algorithms (knn, neural network, etc.).
